---
title: "Packages"
description: |
  Help with installing and understanding packages, at home and in the wild.
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
```


## Terminology

#### Packages and libraries

#### CRAN

#### Functions, arguments, and objects

#### Documentation

## Installing packages

Packages are collections of functions. As we'll see shortly, we use functions as code to view, manipulate, and analyze our data. 

There are packages that come built-in with R. These have names like {base}, {utils}, and {stats}.

Since R is open-source, users are able to create their own packages so that other R users can use them. These packages are available in places like [the Comprehensive R Archive Network (CRAN for short)](https://cran.r-project.org/web/packages/) and [GitHub](https://github.com/).

Lucky for us, packages are easily retrieved from the R Console. If you haven't already, run the following code from your console to download the packages that we'll be needing for Fall Semester. __You only need to do this once__:


```{r install packages, eval = FALSE, echo = TRUE}
install.packages("tidyverse")
install.packages("skimr")
install.packages("epiR")
install.packages("devtools")
devtools::install_github("potato-nathan/epiAssist")
```


## Loading libraries

To enable R to use a package's functions in our current project environment, we need to load the packages using the `library()` function.

In a fresh code chunk, call in the {tidyverse} and {skimr} packages using the following code:

```{r load libraries, eval = FALSE, echo = TRUE}
library(tidyverse)
library(skimr)
```

__Notice that when we install packages, we need to specify their names using quotes, but when we load them into R using `library()`, R recognizes them as package objects automatically, and so don't need quotes. You can surround them with quotes and it will load all the same.__


