---
title: "R Jargon"
description: |
  How to talk like an R nerd. Here, we detail various R-related vocabulary that will make your life a lot easier.  
author:
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(tidyverse)
library(knitr)
```

Does a pesky word keep popping up that's giving you the sneaking suspicion that you have no idea what you're doing? 

I've got some good news, but also some bad news. 

__The bad:__ The person writing this probably has no idea what *they're* doing either. And unless you study computer engineering, there's a good chance you will never really "know" what you are "doing".

__The good:__ A quick vocabulary lesson will help do away with that sneaking suspicion, at least temporarily. 


## R Vocabulary List

Search this document using ctrl+f for a brief definition of that word or symbol and external links for further reading. Can't find the word here? Try R Documentation or Google.

### tibble:

Like a data.frame object, but with enhanced "printing". 

Read [this section on tibbles](https://r4ds.had.co.nz/tibbles.html#tibbles-vs.-data.frame) from R for Data Science to learn more.

### $: 

Used to specify a specific variable within a list-like object (like a data.frame, a tibble, a model, an actual list created with `list()`)

Example:

```{r echo = TRUE, eval = FALSE}
# this code uses head() to view the first six observations for the variable "cyl" in dataset "mtcars"

head(mtcars$cyl)
#> [1] 6 6 4 6 8 6
```

### vector:

A vector is a list of values, all of the same type.

We use `c()` to create a vector, separating items with commas when we specify them individually.

The following are all valid vectors:

```{r echo = TRUE, eval = FALSE}
# a vector of numbers 1, 2, 3:
c(1, 2, 3)
#> [1] 1 2 3

# a vector of numbers 1 through 12, and then 20:
c(1:12, 20)
#> [1]  1  2  3  4  5  6  7  8  9 10 11 12 20

# a vector of logical values:
c(TRUE, FALSE, NA, TRUE)
#> [1] "TRUE"  "FALSE" "TRUE" 

# a vector of 5 values randomly drawn from a uniform distribution with min 0 and max 1:
runif(5)
#> [1] 0.97420369 0.04387835 0.68818071 0.13420150 0.30322080
```

Most operations on vectors apply to each value individually:

```{r echo = TRUE, eval = FALSE}
x <- c(1:5)

x + 2
#> [1] 3 4 5 6 7

x * 2
#> [1]  2  4  6  8  10
```
As such, a data frame is just a __*list of vectors*__ of all the same length. That __*list*__ is what's known formally as a "recursive vector". Lists can contain other lists. 

This is a somewhat complex topic. If you're really hungry for more info, [this chapter](https://r4ds.had.co.nz/vectors.html) in R for Data Science is highly informative, but may be confusing at first for those without any programming background. 

### string/character:

Strings are created with either single quotes or double quotes. It indicates that a value is meant to be read as-is, rather than transformed according to R's computational rules for numbers and logical values.

For example, writing the logical value, TRUE as a string makes it unrecognizable to R as logical:

```{r eval = FALSE, echo = TRUE}
a <- TRUE
isTRUE(a)
#> [1] TRUE

b <- "TRUE"
isTRUE(b)
#> [1] FALSE
```

As simple as it sounds, strings are a topic of mind-numbing complexity, as the underlying encoding of text strings governs the way that code is able to interact with data as well as other code. 

R for Data Science gives [a nice introduction to the mechanics of strings here](https://r4ds.had.co.nz/strings.html#strings), but hints at their wider implications in [its chapter on data importation](https://r4ds.had.co.nz/data-import.html).

Regular expressions (not to be confused with reproducible expressions), or regex, are their own beast, [and may help you understand how databases and search engines work](https://www.regular-expressions.info/)

### factor:

A data type that is the preferred way to store categorical variables in R.

Using `factor()`, you can convert:

* __a character to a factor__: This takes all unique values of a character variable and assigns them an underlying number, or "levels".

For example: 

```{r echo = TRUE, eval = FALSE}
dumplings <- c("momos", "pop tarts", "raviolis", "momos", "empanadas", "pierogis")

factor(dumplings)
#> [1] momos     pop tarts raviolis  momos     empanadas pierogis 
#> Levels: empanadas momos pierogis pop tarts raviolis
```

* __an integer to a factor__: This takes integers and makes the lowest number the base level, and the next highest 1, the next highest 2, etc. You can assign a "label" to each respective level with "labels = c()", with a vector, c(), with the same length as the number of levels.

```{r echo = TRUE, eval = FALSE}
integers <- c(1, 2, 3, 2, 3, 2, 1)

factor(integers,
       labels = c("chicken", "egg", "rooster"))
#> [1] chicken egg     rooster egg     rooster egg     chicken
#> Levels: chicken egg rooster
```
Understanding factors mostly takes time. If you want to speed that up, [here's the R for Data Science chapter on factors](https://r4ds.had.co.nz/factors.html)

In it, they reference a few articles for further reading:

* [Wrangling categorical data in R](https://peerj.com/preprints/3163/)
* [stringsAsFActors: An unauthorized biography](https://simplystatistics.org/2015/07/24/stringsasfactors-an-unauthorized-biography/)
* [stringsAsFactors = <sigh>](https://notstatschat.tumblr.com/post/124987394001/stringsasfactors-sigh) -- this one is kinda funny


### library/packages:

I'll quote from [an answer in StackOverflow](https://stackoverflow.com/questions/26900228/what-is-the-difference-between-a-library-and-a-package-in-r) for this one:

"In R, a package is a collection of R functions, data and compiled code. The location where the packages are stored is called the library."

### directory:

["The working directory of a process is a directory of a hierarchical file system"](https://en.wikipedia.org/wiki/Working_directory)

A __directory__ is any file folder on your computer.

Your __root directory__ is the top-most directory on your computer (in Windows, this is the folder calls "C:", for Mac users, it's usually labelled as "Macintosh HD")

In R, your __working directory__ is typically the folder that contains whatever [R Project](https://dghi-biostat.github.io/biostatlab/help_projects.html) you have open.

Say you have a dataset called "data.rds" in your main working directory. You can import it using any number of functions by referring to its file path as simply "data.rds". 

But say you have that dataset in a series of folders within your working directory. The folders are organized like this, which each subsequent folder inside the last, and the data file in the folder titled "data": 

__*working directory* > try1 > fullAnalysis > data__

The "file path" for referring to your data file from your working directory would be this:

"try1/fullAnalysis/data/data.rds"

### tribble:

Short for "transposed tibble", it's a function that allows you to create a tibble by hand, with the syntax and subsequent output:

```{r echo = TRUE, eval = FALSE}
tribble(
  ~colA, ~colB, ~colC, ~colD,
  "a",   1, "Square", "orange",
  "b",   2, "Circle", "maracuya",
  "c",   3, "Rhombus", "cashew"
)
```

```{r}
x <- tribble(
  ~colA, ~colB, ~colC, ~colD,
  "a",   1, "Square", "orange",
  "b",   2, "Circle", "maracuya",
  "c",   3, "Rhombus", "cashew"
)

kable(x)
```

?tribble in the R Console for more details


### ggplot2:

The "gg" in ggplot2 stands for __"grammar of graphics"__. 

To get a high-level overview of ggplot2 basics, I highly recommend [this introduction](https://r4ds.had.co.nz/data-visualisation.html) to data visualization.

If you're really trying to nerd out, you can access the free full text of [__*ggplot2: Elegant Graphics for Data Analysis*__](https://www.springer.com/gp/book/9780387981413) by Hadley Wickham by looking it up using [the Duke Library search engine](https://quicksearch.library.duke.edu/?q=ggplot2) and logging in with your Duke credentials

### magrittr/pipe/%>%:

Chapter from [R for Data Science](https://r4ds.had.co.nz/pipes.html) gives a nice intro. 

For you nerds, [here's a history of the pipe operator in R](http://adolfoalvarez.cl/plumbers-chains-and-famous-painters-the-history-of-the-pipe-operator-in-r/)

### dplyr:

A package that contains a set of functions that help solve "the most common data manipulation challenges". Main functions include:

* `mutate()`
* `select()`
* `filter()`
* `summarize()` (or `summarise()`)
* `arrange()`

[Here's the chapter](https://r4ds.had.co.nz/transform.html) from R for Data Science

You may also find [this vignette](https://dplyr.tidyverse.org/articles/dplyr.html) helpful

### base R:

These are functions that are a part of the original R programming language, and so do not require a call to a package using `library()`.

[Go here](https://stat.ethz.ch/R-manual/R-devel/library/base/html/00Index.html) to see a complete list of functions that come with the R Base Package

### vignette: 

A vignette is a long-form guide to a package. It highlights a package's main functions and their usage. Learning from vignettes is one of the best ways to self-teach yourself a skill in R.

### #: 



### YAML:

### index:

### tidy:

### tidyverse:

### knit:

### console:

### git:

### code chunk:

### traceback:

### html:

### debug:

### function:

### argument:

### parameter:

### reproducible expressions (reprex):

### Hadley Wickham:

### runif/qunif/punif/dunif

