---
title: "Lab 2"
description: |
  Biostatistics & Epi, Lab 1 - GLHLTH 705.
date: "`r Sys.Date()`"
output: distill::distill_article
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(knitr)
```

### LAB MATERIALS: 

* [__R Markdown file for Lab 1.__](assignments/lab1_705_fall2021.Rmd)
Click link to download. Fill it in with your answers to the following lab tasks. When you're ready to submit, name it as `Lab1_FirstinitialYourlastname.Rmd`, and submit it using the Sakai dropbox.

* [__Excel document for Tables 1 & 2__](assignments/Lab1_Tables.xlsx)
TODO: Set up answer entry on Sakai?

* Lab_2_kenya.rds - data file available on Sakai

# Lab 2 Goals:

* Generate derived variables
* Identify and recode special values 
* Run descriptive statistics for continuous and categorical variables 
* Generate graphics for continuous variables 
* Generate a complete data dictionary


# Lab 2 {.tabset}

## Task 1: Load libraries and data

For this assignment, use the dataset that I have supplied you – lab_2_kenya.rds. You will need the {tidyverse} package as well.


```{r}
# load tidyverse library

```


```{r}
# import dataset and assign a name

```


## Task 2: Code derived variables

Construct the following derived variables.  For each, examine the component variables for missing values (e.g. ‘missing’, ‘unknown’) and be sure to set derived variable values appropriately (e.g., check to make sure that your NA values have not been mistakenly added to another category). Variables should be of the correct type. Convert all categorical variables to factors, labeling them appropriately.


### Task 2a: size 

size: Categorical variable describing size of child at birth (subjectively described by mother). Note that greater values indicate smaller size: [1=very large, 2=larger than average, 3=average, 4=smaller than average, 5=very small]. Based on variable m18.

```{r}


```



### Task 2b: belowavg 

belowavg: Dichotomous variable indicating if the child’s size was below average [0= average, larger than average or very large, 1=smaller than average, very small]

```{r}

```

```{r}

```


### Task 2c: pnc 

pnc: Dichotomous categorical variable for any prenatal care [0 = the referent category; no prenatal care; 1 = the index category; received prenatal care]. Based on variable m2n.

```{r}

```




### Task 2d: rural 

rural: Dichotomous indicator of rural residence [0=urban, 1=rural]. Based on variable v025.

```{r}

```




### Task 2e: education 

education: Categorical educational level obtained by mother [create values for 0=did not attend school, 1=primary school only, 2=post-primary education]. Based on variable s109.

```{r}

```

```{r}

```



### Task 2f: death 

death: Dichotomous categorical variable for death within the first 5 years of life [0 = child alive at 5th birthday (the referent category); 1 = child died before 5th birthday (the index category)].  Based on the variable b5 (death indicator). 

```{r}

```

```{r}

```



### Task 2g: time 

time: Continuous variable for the age at death OR age at interview for children still alive. Based on b7 (age at death for children dead at interview); for children alive at time of interview, calculate their age using v008 (date of interview) and b3 (date of birth). Note that times are given in months.

```{r}

```




```{r}

```



## Task 3: Close the cohort

Create the equivalent of a closed cohort for the analysis of 5-year childhood mortality from this dataset  . In a closed cohort, everyone must be at risk of the outcome at entry into the cohort, and all members of the cohort must remain at risk until they experience the outcome or complete the entire risk period for the outcome. A closed cohort allows for estimation of absolute (unconditional) risks and associated effect measures. For this study, the risk period for child mortality begins at birth and ends when a child dies, or when 60 months of life have been completed. I recommend completing the steps below in order to generate what we want. There are other orders in which to do this, but sometimes you will get incorrect results. If you’re adventurous, I recommend trying other ways to code this; you’ll learn a lot from that exercise.

*__NOTE:__* Complete task 3 before completing the remaining tasks.

### Task 3a: Exclude irrelevant observations


Exclude those observations from the dataset where the child was alive (death=0) and  follow-up time (time) was less than 60 months (think about which boolean operator you’ll need to exclude those children). You should end up with 16,828 subjects in the dataset. In this step, you should also use an assignment operator (<-) to assign this closed cohort to a different object name. That way, you will be able to work with this dataset separately in the coming steps.


```{r}

```


Task 3b: 

Finally, recode the time variable to indicate the month of life during which death or interview occurred (add 1 month to the current value). The range on time should now be 1 month – 61 months. 

```{r}

```




Task 3c: Recode time +1

Finally, recode the time variable to indicate the month of life during which death or interview occurred (add 1 month to the current value). The range on time should now be 1 month – 61 months. 

```{r}

```


### Task 3e: Save new dataset

Save this new dataset containing 16828 subjects (make sure to give it a new name).

```{r}

```


## Task 4: Frequency histograms

Generate frequency histograms of time, overall and separately for levels of death.  Put meaningful axis labels and a title on each figure.

```{r}

```


```{r}

```

```{r}

```


```{r}

```



## Task 5: Boxplots of time

Generate boxplots of time, overall and separately for levels of death.  Put meaningful axis labels and a title on each figure.

```{r}

```


```{r}

```

```{r}

```


## Task 6: Frequency counts and percentages

Fill in Table 1 with the frequency counts and percentages for the levels of the new categorical variables you have generated.  Calculate percentages only for the non-missing values.  Round percentages to 1 decimal place. 

### Size at birth (size)

```{r}

```

### Size at birth categorical (belowavg)
```{r}

```

### Prenatal care (pnc)
```{r}

```

### Residence type (rural)
```{r}

```
### Mother's education (education)
```{r}

```

### Death by 5 years (death)

```{r}

```


## Task 7: Update data dictionary 

Update your data dictionary for the Kenya dataset, adding the newly created variables.

```{r}

```


## Task 8: Short Answer

__Examine the range and proportion of missing for each of the variables 7 you have created in this lab.  Are there characteristics of any of these variables that are concerning (e.g., missing, suspicious or impossible values)?  In contemplating analysis of these data, what do you think should be done with anomalous information?  What effect would missing values have on the validity of your analyses (e.g., how might missing or extreme values affect inferences)?__


