---
title: "GLHLTH 705 Lab"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
---

# Lab 0 {.tabset}

## Task 3: Install packages and load libraries

__a. Note that for this step, you should first install three packages using the following codes; If you have already installed these packages once before, proceed to task 3b directly.__

`install.packages(c("tidyverse", "skimr", "epiR", "tableone", "devtools"))`
`devtools::install_github("potato-nathan/epiAssist")`

__b. Load libraries:__

__For this lab, we need {skimr} and {tidyverse}__


__In a fresh code chunk, call in the {tidyverse} and {skimr} packages using the following code:__

`library(tidyverse)`
`library(skimr)`

```{r load libraries}
# load library {tidyverse}


# now load library {skimr}

```


## Task 4: Load data

__Use command readRDS(‘filename.rds’) to load in the dataset. Don’t forget to use the assignment operator to give the dataset a name after importing the data. As an example, you can name it like this:__

`kenya <- readRDS('data/filename.rds')` 

```{r load data}
# load dataset and give it a name within the r environment


```



## Task 5: Explore the dataframe

__Familiarize yourself with the data by using the commands `head()`, `skim()`, `ncol()`, and `nrow()`.__

* Are there any string/character variables?
* Are there any variable or value labels? 
* Do any variables have notes? 

```{r dimensions}
# find number of columns with ncol(dataName)

# find number of rows with nrow(dataName)
```

```{r variables}
# inspect variable names with names(dataName)

# inspect 3 variable's types with class(dataName$variableName)

```

```{r skim}
# generate summary statistics for your dataset with skim(dataName)

```



## Task 6: Create variable `mage`

__Using a pipe and the `mutate()` function, create a new variable, `mage` for mother’s age (as an integer) at the time of each child’s birth (note – some of these mothers have had multiple children). __

__This is calculated from variables `b3` (month code of child’s birth) and `v011` (month code of mother’s birth). The difference between the values of these variables is in months, `so divide by 12 to get years`. See the data dictionary for a more detailed description of month codes and how to use them. Use `as.integer()` around your calculation to truncate the calculated values for `mage` to integers.__


```{r mage}
# create variable mage with a pipe (%>%) and mutate()


```

## Task 7: Frequency distributions of mage

__Suppose you want to break down `mage` into an ordinal categorical variable with three categories. First, we might inspect the frequency distribution (in one-way frequency table) for `mage`. Do this using `table()`.__

__Use `table()` to look at `kenya$mage`, then consider the following questions:__

* Are there any missing values for `mage`? If so, how many?  
* Which range of ages appear the most frequently in `mage`?  

```{r mage table}
# generate a table of mage's frequency distribution with table()
# don't forget to include counts of NA values with the argument useNA = "always"


```


## Task 8: Create variable magec

__Using `mage`, generate a new variable with three categories: <18, 18-39, and ≥ 40, naming the new variable "magec” (mother’s age categorical). Set the values for `magec` to be 0,1,2, where 0 corresponds to the youngest age group (<18).__
__Recommended steps:__

```{r create magec}
# create magec with a pipe (%>%) and mutate()
# since this is your first time encountering case_when(), we've provided a layman's translation of how our code is communicating with the dataset. You just need to specify x, y, and z accordingly:


kenya <- kenya %>%
  mutate(magec = 
           case_when(mage < 18 ~ x, 
                     # "if mage is less than 18, then magec is x"
                     mage >= 18 & mage < 40 ~ y, 
                     # "if mage is greater than or equal to 18 AND if it is less than 40, then magec is y
                     mage >= 40 ~ z) 
                     # " if mage is greater than or equal to 40, then magec is z 

  ) # don't forget the closing parenthesis for mutate()

# finally, notice that each new condition is separated from the last by a comma

```


```{r factor magec}
# factor magec (convert it to a categorical variable) with factor()

```



## Task 9: Cross-tab of mage and magec

__Look at a cross-tabulation (two-way table) of mage and magec to ensure that magec was created correctly. Be sure missing values were handled properly (all observations that have a missing value for mage should be assigned the R missing value “NA” for magec.).   Type “?table” for help with how to create a 2x2 table.  Note: the order of the variables in the command controls which one is in the rows and which is in the columns of the table. Experiment to make your table readable.__

### Method 1:

```{r table cross-tab}
# create a cross-tabulation of your new variables with table()

```

### Method 2: 

```{r tidyverse cross-tab}
# generate cross-tabulated numbers with a pipe, group_by(), and count()

```




## Task 10: Save new dataset

__Save the new dataset in the same directory using a different file name with the form shown in the instructions on page 1 using the saveRDS() command.__

```{r save data}
# save the datset with saveRDS()

saveRDS(datasetName, file = "data/filename.rds")
```








