---
title: "Lab 1 - Kenya"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: html_document
---

# Lab 1: {.tabset}

## Goals

* Generate derived variables
* Identify and recode special values 
* Run descriptive statistics for continuous and categorical variables 
* Generate graphics for continuous variables 
* Generate a complete data dictionary

## Task 1: Load libraries & packages

Use the dataset Lab_1_kenya.rds for this assignment. You will also need packages {skimr} and {tidyverse}

```{r}
# load libraries

Sys.
```

```{r}
# load data

```


## Task 2: Recode variables

Construct the following derived variables using a pipe and `mutate()`. For each, examine the component variables for coded special values (e.g., missing, unknown) and be sure to set derived variable values appropriately.  Label all variables and the coded values for the categorical ones using the `factor()` command.

### Task2a: bord5

bord5: Dichotomous variable indicating birth order of the child. [0=current child was 1st through 4th in the birth order, 1=current child was later than 4th in the birth order]. Based on variable `bord` (birth order).


```{r}
# create derived variable using %>% and mutate()


# label derived variable with factor()


```


### Task 2b: male

male: Dichotomous categorical variable indicating that the child is male [0=female, 1=male]. Based on variable `b4`.

```{r}
# create derived variable using %>% and mutate()


# label derived variable with factor()


```

### Task 2c: mweight

mweight: Continuous variable for maternal weight at time of interview (in kilograms). Based on variable `v437`. Note that `v437` contains 1 implied decimal place. Divide by 10 to get kilograms. 

```{r}
# create derived variable using %>% and mutate()


# label derived variable with factor()


```

### Task 2d: mheight

mheight:  Continuous maternal height at time of interview (in meters). Based on variable `v438`. Note that this variable is in centimeters and also contains 1 implied decimal place.

```{r}
# create derived variable using %>% and mutate()


# label derived variable with factor()


```


### Task 2e: mbmi

mbmi: maternal body mass index (BMI).  Weight (kilograms) / height^2 (meters). 

```{r}
# create derived variable using %>% and mutate()


# label derived variable with factor()


```


## Task 3: Frequency histograms

Generate frequency histograms of mweight, mheight and mbmi.  Put meaningful axis labels and a title on each figure.


### Task 3a: mweight

```{r}

```


```{r}

```

### Task 3b: mheight

```{r}

```

### Task 3c: mbmi

```{r}

```

## Task 4: Boxplots

Generate boxplots of mweight, mheight, and mbmi for the levels of magec. Put meaningful axis labels and a title on each figure.

### Task 4a: mweight

```{r}

```

### Task 4b: mheight

```{r}

```

### Task 4c: mbmi

```{r}

```


## Task 5: Frequency table

Fill in Table 1 with the frequency counts and percentages for the levels of the 3 categorical variables you have generated. Calculate percentages only for the non-missing values. Round percentages to 1 decimal place.

### Task 5a: bord5

```{r}
# frequency counts


# frequency percentages

```

### Task 5b: magec
```{r}
# frequency counts


# frequency percentages

```

### Task 5c: male

```{r}
# frequency counts


# frequency percentages

```

## Task 6: Table 2

Table 2 is the data dictionary for the Kenya dataset, with columns added to annotate the variables and provide summary statistics.

a.	Fill in these columns in this table for the variables originally on the dataset.
b.	Add rows for the 7 new variables that you have created in both Labs 0 and 1.

```{r}

```

## Task 7: Short answer

__Examine the range and proportion of missing values for each of the 7 variables you have created in Labs 0 and 1. Are there characteristics of any of these variables that are concerning (e.g., missing, suspicious or impossible values)? In contemplating analysis of these data, what do you think should be done with anomalous information? What effect would missing values have on the validity of your analyses (e.g., how might missing or extreme values affect inferences)?__
__(Response no longer than 250 words, please)__


(Type your response here! No need to put it in a code chunk.)

