---
title: "GLHLTH 705 Lab"
subtitle: "Due: Friday November 12, 2021"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
---


# Lab 3 {.tabset}

## Task 1: Load libraries and dataset

```{r}
# load the {tidyverse}, {tableone}, {epiR}, and {epiAssist} packages

```

```{r}
# load in the dataset

```

## Task 2: BIG PICTURE 

#### Just think on this for a moment. Internalize it. Come back to it:

For this and subsequent labs, we are considering 1) child mortality within 5 years as our outcome (response) variable, 2) birth order >= 5 as the main risk factor variable and 3) other variables as potential confounders or modifiers of the relationship between mortality and birth order. 


## Task 3: Table 1 

### Tabular analysis of continuous variables by death


### Maternal BMI by death

```{r}
# use a pipe and skim() to get maternal BMI overall

```


```{r}
# use a pipe with group_by() and then skim() to get summary stats of mage, stratified by `death`


```


```{r}
# use a pipe with group_by() and then count() to get frequency counts for `death`


```


```{r}
# T-test for maternal bmi and death using t.test()

```

```{r}
# Kruskal-Wallis test for maternal bmi and death using kruskal.test()


```


### Maternal age (continuous) by death

```{r}
# use a pipe and skim() to get maternal age overall

```

```{r}
# use a pipe with group_by() and then skim() to get summary stats of mage, stratified by `death`


```

```{r}
# T-test for maternal age and death using t.test()

```

```{r}
# Kruskal-Wallis test for maternal age and death using kruskal.test()


```



## Task 4: Table 2

### Tabular analysis of categorical variables by death


```{r}
# create a tableone object with CreateTableOne() for overall counts

```

```{r}
# create a tableone object, this time specifying a value for "strata = "
```

```{r}
# to perform a fisher's exact test, create one last tableone object, then submit it to its print() function, with the argument "exact = TRUE"


```


## Task 5: Table 3

Fill in Table 3 with frequency counts overall and by levels of child mortality within 5 years and with the stratum-specific risk of death, the risk difference, risk ratio and the odds ratio of death comparing levels for each of the 5 variables to the reference group.  The reference group for each variable is noted.  For the two 3-level variables, calculate risk differences and ratios for each non-referent level relative to the reference group.  Round risks to 3 decimal places and ratios to 2 decimal places.  The STATA cs command provides a convenient way to produce the data for this table.

### Variable: bord5

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```

### Variable: male

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```


### Variable: rural

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```

### Variable: magec

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```



### Variable: education

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```

## Task 6: Side-by-side boxplots

### mbmi by death
```{r}
# use ggplot to stratify boxplots by using a categorical variable on either the x or y axis

# save your plot with ggsave()

```

### mbmi by death
```{r}
# use ggplot to stratify boxplots by using a categorical variable on either the x or y axis

# save your plot with ggsave()

```



## APPENDIX

* {tableone} - I've used this package a LOT for generating data reports. It is a super convenient tool for compiling baseline data. It's also easy to copy/paste into excel using the argument quote = TRUE, which creates character-delineation between values, which excel recognizes as columns. [Find a tutorial on all of that here](https://cran.r-project.org/web/packages/tableone/vignettes/introduction.html)



