---
title: "GLHLTH 705 Lab"
subtitle: "Due: Friday November 12, 2021"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
---


# Lab 3 {.tabset}

## Task 1: Load libraries and dataset

```{r}
# load the {tidyverse}, {tableone}, {epiR}, {skimr}, and {epiAssist} packages

```

```{r}
# load in the dataset

```

## Task 2: BIG PICTURE 

#### Just think on this for a moment. Internalize it. Come back to it:

For this and subsequent labs, we are considering 1) child mortality within 5 years as our outcome (response) variable, 2) birth order >= 5 as the main risk factor variable and 3) other variables as potential confounders or modifiers of the relationship between mortality and birth order. 


## Task 3: Table 1 

### Tabular analysis of continuous variables by death


### Maternal BMI by death

```{r}
# use a pipe and skim() to get maternal BMI overall

```


```{r}
# use a pipe with group_by() and then skim() to get summary stats of mage, stratified by `death`


```


```{r}
# use a pipe with group_by() and then count() to get frequency counts for `death`


```


```{r}
# T-test for maternal bmi and death using t.test()

```

```{r}
# Kruskal-Wallis test for maternal bmi and death using kruskal.test()


```


### Maternal age (continuous) by death

```{r}
# use a pipe and skim() to get maternal age overall

```

```{r}
# use a pipe with group_by() and then skim() to get summary stats of mage, stratified by `death`


```

```{r}
# T-test for maternal age and death using t.test()

```

```{r}
# Kruskal-Wallis test for maternal age and death using kruskal.test()


```



## Task 4: Table 2

### Tabular analysis of categorical variables by death


```{r}
# create a tableone object with CreateTableOne() for overall counts

```

```{r}
# create a tableone object, this time specifying a value for "strata = "
```

```{r}
# to perform a fisher's exact test, create one last tableone object, then submit it to its print() function, with the argument "exact = TRUE"


```


## Task 5: Table 3

Fill in Table 3 with frequency counts overall and by levels of child mortality within 5 years and with the stratum-specific risk of death, the risk difference, risk ratio and the odds ratio of death comparing levels for each of the 5 variables to the reference group.  The reference group for each variable is noted.  For the two 3-level variables, calculate risk differences and ratios for each non-referent level relative to the reference group.  Round risks to 3 decimal places and ratios to 2 decimal places.  The STATA cs command provides a convenient way to produce the data for this table.

### Variable: bord5

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```

### Variable: male

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```


### Variable: rural

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```

### Variable: magec

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```



### Variable: education

```{r}
# create 2x2 table object


# flip table with flipTable()


```


```{r}
# get measures of association with mAssoc()


```

## Task 6: Side-by-side boxplots

### mbmi by death
```{r}
# use ggplot to stratify boxplots by using a categorical variable on either the x or y axis

# save your plot with ggsave()

```

### mage by death
```{r}
# use ggplot to stratify boxplots by using a categorical variable on either the x or y axis

# save your plot with ggsave()

```


## Task 7: Short answer

For __Table 1__, what is the null hypothesis being tested for each the two variables.  Do the two statistical tests agree?  Is there a statistically significant difference in the distributions of either BMI or age by child mortality?  Are there substantively/clinically significant differences in these distributions?  If statistical and substantive differences disagree, why might that be the case?


## Task 8: Short answer
For __Table 2__, what is the null hypothesis being tested for the variables.  Do the two statistical tests agree?  Is there a statistically significant difference in any of these bivariate distributions?  Are there substantively/clinically significant differences in these distributions?  

## Task 9: Short answer
Age has been evaluated both as a continuous and categorical variable.  Do the tests of association for these 2 parameterizations of age agree?  If not, why not?


## Task 10: Short answer
For __Table 3__, which risk factor has the strongest association with mortality?  Provide a rationale for this assessment.  Which factors are protective and which are harmful?


## Task 11: Short answer
Prepare a brief paragraph describing and interpreting risk, RD and RR estimates (from __Table 3__) for the 60 month risk of death in association with birth order and maternal age. 

Begin the paragraph by reporting the overall risk or proportion of deaths in the study population. Don’t just repeat numbers that are already shown in the table; note the values of effect estimates you feel should be highlighted, but otherwise describe the results in words. Write this paragraph as if you were including it in the results section of a manuscript. (175 words maximum).


