---
title: "GLHLTH 705 Lab"
subtitle: "Due: Friday November 23, 2021"
author: "YOUR NAME HERE"
date: "`r Sys.Date()`"
output: 
  html_document:
    df_print: paged
---

# Lab 4 {.tabset}

## Task 1: Load libraries & data

For this assignment, use the dataset 'Lab_4_kenya.rds'.

```{r}
# load libraries: {tidyverse}, {epiR}, {fmsb}, {skimr}, and {epiAssist}

```


```{r}
# load in dataset 

```

## Task 2: Tables 1 & 2 {.tabset .tabset-pills}

Fill in tables 1 and 2 with the numbers of deaths and the 60-month risk of death according to birth order (bord5) stratified by:

a)	Child’s gender (variable male)
b)	Rural/urban residence (variable rural)
c)	Maternal age (variable magec)
d)	Maternal education (variable education)
e)	Calculate the RD (Table 1) and RR (Table 2) for mortality in association with bord5 within each covariable stratum
f)	Calculate a homogeneity test statistic and p-value for each covariate listed above
g)	To get the stratified and pooled estimates and p-values, run the CS command, and you will need to use the estandard and pool and rd options for RD and the pool option for the RR.
h)	For each covariable listed above, report the E. Standardized pooled estimate estimate for the RD and the Mantel-Haenszel pooled estimate (M-H combined) for the RR


### Task 2.1: Frequency counts and risk

#### 2.1.a: `male`

```{r}
# use dplyr functions to calculate sub-group frequency counts and risk of death by sex (male) and birth order (bord5)

  
```

#### 2.1.b: `rural`

```{r}
# use dplyr functions to calculate sub-group frequency counts and risk of death by residence (rural) and birth order (bord5)

  
```

#### 2.1.c: `magec`

```{r}
# use dplyr functions to calculate sub-group frequency counts and risk of death by mother's age (magec) and birth order (bord5)

  
```

#### 2.1.d: `education`

```{r}
# use dplyr functions to calculate sub-group frequency counts and risk of death by mother's level of education (education) and birth order (bord5)

  
```

### Task 2.2: Create `epi.2by2()` object (TODO: epi.2by2() or mAssoc()?)

#### 2.2.a: Create table object with `table()`

```{r}
# male table

# rural table

# magec table

# education table
```


#### 2.2.b: Flip table with `flipTable()`

```{r}
# flip male

# flip rural

# flip magec

# flip education

```


#### 2.2.c: Create epi.2by2 object with `epi.2by2()`

```{r}
# epi.2by2() for male

# epi.2by2() for rural

# epi.2by2() for magec

# epi.2by2() for education

```

### Task 2.3: Risk differences and test of homogeneity (TODO: how to get risk difference? seems borken at the moment in epi.2by2())

#### 2.3.a: Extract risk differences from within epi.2by2() object

```{r}
# Risk difference:

## male

## rural

## magec

## education

```

#### 2.3.b: Extract homogeneity test statistics

```{r}
# Homogeneity test statistic and p value

## male

## rural

## magec

## education
```

#### 2.3.b: Report standardized pooled estimate

```{r}
# standardized pooled estimate

## male

## rural

## magec

## education
```


### Task 2.4: Risk ratios and test of homogeneity

#### 2.4.a: Extract risk ratios from within epi.2by2() object

```{r}
# Risk difference:

## male

## rural

## magec

## education

```

#### 2.4.b: Extract homogeneity test statistics

```{r}
# Homogeneity test statistic and p value

## male

## rural

## magec

## education
```

#### 2.4.c: Report Mantel-Haenszel pooled estimate (M-H combined)

```{r}
# Mantel-Haenszel pooled estimate

## male

## rural

## magec

## education
```






