---
title: "mantel-haenszel"
author: "DGHI Biostat"
date: "11/6/2021"
output: html_document
---

## Toy data
```{r}

# list of cell values
num <- c(23, 53, 46, 23, 20, 68, 58, 34)

# create a 3-D array and convert to a table object
dat <- as.table(array(num, dim = c(2,2,2)))

# voila!
dat
```


## Arrange data

Vectors for each cell position
(i.e. `a` will be a vector that contains values from that top-left cell of all strata)
```{r}
  if(length(dim(dat)) > 2){
    a <- dat[1,1,]; A <- a
    b <- dat[1,2,]; B <- b
    c <- dat[2,1,]; C <- c
    d <- dat[2,2,]; D <- d
  }

# compare a with printed table in previous chunk
a
```



Layout of 2x2 contingency tables

```{r}

  ## =================
  ## DECLARE VARIABLES
  ## =================
  
  ##        | D+   | D-   | Total
  ## ----------------------------
  ## Exp +  | a    | b    | N1
  ## Exp -  | c    | d    | N0
  ## -------|------|------|------
  ## Total  | M1   | M0   | Total

```


Totals within exposure strata and outcome (case) strata
```{r}
# Total within strata cases:
  M1 <- a + c
  # Total within strata non-cases:
  M0 <- b + d
  # Total within strata exposed:
  N1 <- a + b
  # Total within strata unexposed:
  N0 <- c + d
  # Total within strata subjects:
  total <- a + b + c + d
```

## Calculate crude estimates

Risk difference
```{r}
crude_RD <- (a / (a + b)) - (c / (c + d))
crude_RD
```

Risk ratio
```{r}
crude_RR <- (a / (a + b)) / (c / (c + d))
crude_RR
```



## Calculate MH pooled estimates

### Pooled risk difference (M-H)
(note: I also found this calculation being referred to as a "Mantel-Haenszel type method". source: https://www.statsdirect.com/help/meta_analysis/risk_difference.htm)
```{r}
# assign units (units = 1 provides calculation of risk difference)
units <- 1

  # Summary attributable risk (Rothman 2002 p 147 and p 152, equation 8-1):
sARisk.p <- (sum(((a * N0) - (c * N1)) / total) / sum((N1 * N0) / total)) * units
  # w <- (N1 * N0) / total
  # var.p1 <- (((a * d) / (N1^2 * (N1 - 1))) + ((c * b) / (N0^2 * (N0 - 1))))
  # var.p1[N0 == 1] <- 0
  # var.p1[N1 == 1] <- 0
  # varARisk.s <- sum(w^2 * var.p1) / sum(w)^2
  # sARisk.se <- (sqrt(varARisk.s)) * units
  # sARisk.l <- sARisk.p - (z * sARisk.se)
  # sARisk.u <- sARisk.p + (z * sARisk.se)

# check point estimate against crude estimates

sARisk.p
crude_RD
```


### Pooled incidence risk ratio (M-H)
(note: this one is for sure an M-H method. source: https://sphweb.bumc.bu.edu/otlt/mph-modules/bs/bs704-ep713_confounding-em/BS704-EP713_Confounding-EM7.html)
```{r}
  # Summary incidence risk ratio (Rothman 2002 p 148 and 152, equation 8-2):
  sRR.p <- sum((a * N0 / total)) / sum((c * N1 / total))
  # varLNRR.s <- sum(((M1 * N1 * N0) / total^2) - ((a * c)/ total)) / (sum((a * N0)/total) * sum((c * N1)/total))
  # lnRR.s <- log(sRR.p)
  # sRR.se <- (sqrt(varLNRR.s))
  # sRR.l <- exp(lnRR.s - (z * sqrt(varLNRR.s)))
  # sRR.u <- exp(lnRR.s + (z * sqrt(varLNRR.s)))

# check point estimate
sRR.p
crude_RR
```



## Reference

Source code: https://rdrr.io/cran/epiR/src/R/epi.2by2.R

Internal citation: Rothman KJ (2002). Epidemiology An Introduction. Oxford University Press, London, p 130 - 143.
